---
title: "Gayane"
output: html_notebook
---

Date: 11/18/2023

Primary Hypothesis:   
Correlate baseline_D-wave_distal decrease with low parameters in clinical data (scores etc).    

# Setup
```{r}
require(data.table)
require(ggplot2)
require(table1)
require(formatters)
```

# Data
## Load data
```{r}
dt_clin <- fread("Combined data Slutgiltig Clinical data 11182023.csv")
dt_np <- fread("Combined data Slutgiltig Neurophysiology data 11182023.csv")
```

## Remove Subject 53 and Subject 62 second surgeries (more than 10 years later)
```{r}
dt_np <- dt_np[!(`Study nr` == 53 &
                   Op_datum == "4/13/2021"), ]

dt_np <- dt_np[!(`Study nr` == 62 &
                   Op_datum == "12/10/2019"), ]
```

## Merge
```{r}
dt1 <- merge(dt_clin,
             dt_np,
             by = "Study nr",
             all = TRUE)
```

## Clear
```{r}
rm(list = c("dt_clin",
            "dt_np"))
gc()
```

## Format outcomes
```{r}
# Intraoperative D-wave distal
# "Not detected" = 100% loss of D-wave
# N/A = not measured
# "" = remove 1 patient (Study nr = 5) as no data is available for them
dt1 <- dt1[`Intraoperative_D-wave_distal` != "",]

# Create new variable for the analysis, merging D-wave groups
dt1$interop_d_distal <- unlist(lapply(X = substr(x = tolower(dt1$`Intraoperative_D-wave_distal`),
                                                 start = 1,
                                                 stop = 3),
                                      FUN = function(a) {switch(a,
                                                                "unc" = "unchanged",
                                                                "non" = "non-signif decrease",
                                                                "sig" = "signif decrease",
                                                                "not" = "not detected",
                                                                "abs" = "not detected",
                                                                "not measured/unavailable")}))

dt1$interop_d_distal <- factor(dt1$interop_d_distal,
                               levels = c("unchanged",
                                          "non-signif decrease",
                                          "signif decrease",
                                          "not detected",
                                          "not measured/unavailable"))

# Baseline D-wave distal
dt1$`baseline_D-wave_distal`[dt1$`baseline_D-wave_distal` == ""] <- "normal"
dt1$baseline_d_distal <- factor(dt1$`baseline_D-wave_distal`,
                                levels = c("normal",
                                           "low amplitude",
                                           "not detected"))

# Table
addmargins(table(interop_d_distal = dt1$interop_d_distal,
                 baseline_d_distal = dt1$baseline_d_distal))
```

## Errata
### Typos
```{r}
dt1$`Long-term pain change`[dt1$`Long-term pain change` == "wose"] <- "worse"

dt1$Intraoperative_SEP_left_foot[dt1$Intraoperative_SEP_left_foot == "ampitude decrease"] <- "amplitude decrease"
```

### Shorthand
```{r}
dt1$`Long-term bowel change`[dt1$`Long-term bowel change` == "u"] <- "unchanged"
```

### Baseline BCR
```{r}
dt1$baseline_BCR[dt1$baseline_BCR == ""] <- "normal"

# Study nr = 10: missing at baseline, intraoperative = N/A
dt1$baseline_BCR[dt1$`Study nr` %in% c(10)] <- "N/A"
```

## Format other variables
```{r}
dt2 <- data.table(study_id = dt1$`Study nr`,
                  
                  # D-wave
                  # dt1$`baseline_D-wave_proximal`, # All NA
                  # dt1$`Intraoperative_D-wave_proximal` # All `unchanged` or NA
                  
                  d_wave_distal_baseline = dt1$baseline_d_distal,
                  d_wave_distal_intraoperative = dt1$interop_d_distal,
                  
                  surgery_date = as.Date(x = dt1$`date surgery`,
                                         format = "%Y/%m/%d"),
                  op_date = as.Date(dt1$Op_datum,
                                    format = "%Y/%m/%d"),
                  long_term_followup_date = as.Date(x = dt1$`Long-term follow-up date`,
                                                    format = "%Y/%m/%d"),
                  
                  # Death
                  dead = dt1$dead == 1,
                  dead_spinal_pathology = dt1$`Death due to spinal pathology` == 1,
                  death_date = as.Date(x = dt1$`date death`,
                                       format = "%Y/%m/%d"),
                  
                  sex = factor(dt1$male,
                               levels = 0:1,
                               labels = c("Female",
                                          "Male")),
                  age = dt1$age,
                  
                  # Tumor
                  pad = factor(dt1$pad),
                  myelomalacia  = factor(dt1$myelomalacia,
                                         levels = 0:1,
                                         labels = c("No",
                                                    "Yes")),
                  contrast_enhancing_tumor = factor(dt1$`contrast enhancing tumor`,
                                                    levels = 0:1,
                                                    labels = c("No",
                                                               "Yes")),
                  syrinx = factor(dt1$syrinx,
                                  levels = 0:1,
                                  labels = c("No",
                                             "Yes")),
                  
                  who = as.numeric(dt1$who),
                  symptom_duration_mo = as.numeric(dt1$`symptom duration (months)`),
                  
                  # Motor
                  motor_pre = factor(dt1$`pre motor`,
                                     levels = 0:1,
                                     labels = c("No",
                                                "Yes")),
                  motor_3m = factor(dt1$`3m motor`,
                                    levels = 0:1,
                                    labels = c("No",
                                               "Yes")),
                  motor_3m_change = factor(tolower(dt1$`3m  motor change`),
                                           levels = c("unchanged",
                                                      "worse",
                                                      "better")),
                  motor_long_term = factor(dt1$`Long-term motor`,
                                           levels = c("0",
                                                      "1",
                                                      "N/A"),
                                           labels = c("No",
                                                      "Yes",
                                                      "n/a")),
                  motor_long_term_change = factor(tolower(dt1$`Long-term motor change`),
                                                  levels = c("unchanged",
                                                             "worse",
                                                             "better",
                                                             "n/a")),
                  
                  # Sensory
                  sensory_pre = factor(dt1$`pre sensoty`,
                                       levels = 0:1,
                                       labels = c("No",
                                                  "Yes")),
                  
                  sensory_3m = factor(dt1$`3m sensory`,
                                      levels = c("0",
                                                 "1"),
                                      labels = c("No",
                                                 "Yes")),
                  sensory_3m_change = factor(tolower(dt1$`3m sensory change`),
                                             levels = c("unchanged",
                                                        "worse",
                                                        "better")),
                  sensory_long_term = factor(dt1$`Long-term sensory`),
                  sensory_long_term_change = factor(tolower(dt1$`Long-term sensory change`),
                                                    levels = c("unchanged",
                                                               "worse",
                                                               "better",
                                                               "n/a")),
                  # Pain
                  pain_pre = factor(dt1$`pre pain`,
                                    levels = 0:1,
                                    labels = c("No",
                                               "Yes")),
                  pain_3m = factor(dt1$`3m pain`,
                                   levels = 0:1,
                                   labels = c("No",
                                              "Yes")),
                  pain_3m_change = factor(tolower(dt1$`3m pain change`),
                                          levels = c("unchanged",
                                                     "worse",
                                                     "better")),
                  pain_long_term = factor(dt1$`Long-term pain`,
                                          levels = 0:1,
                                          labels = c("No",
                                                     "Yes")),
                  pain_long_term_change = factor(tolower(dt1$`Long-term pain change`),
                                                 levels = c("unchanged",
                                                            "worse",
                                                            "better")),
                  
                  # Bowel
                  bowel_pre = factor(dt1$`pre bowel`,
                                     levels = 0:1,
                                     labels = c("No",
                                                "Yes")),
                  bowel_3m = factor(dt1$`3m bowel`,
                                    levels = 0:1,
                                    labels = c("No",
                                               "Yes")),
                  bowel_3m_change = factor(tolower(dt1$`3m bowel change`),
                                           levels = c("unchanged",
                                                      "worse",
                                                      "better")),
                  bowel_long_term = factor(dt1$`Long-term bowel`,
                                           levels = 0:1,
                                           labels = c("No",
                                                      "Yes")),
                  bowel_long_term_change = factor(tolower(dt1$`Long-term bowel change`),
                                                  levels = c("unchanged",
                                                             "worse",
                                                             "better")),
                  
                  # Bladder
                  bladder_pre = factor(dt1$`pre bladder`,
                                       levels = 0:1,
                                       labels = c("No",
                                                  "Yes")),
                  bladder_3m = factor(dt1$`3m bladder`,
                                      levels = 0:1,
                                      labels = c("No",
                                                 "Yes")),
                  bladder_3m_change = factor(tolower(dt1$`3m bladder change`),
                                             levels = c("unchanged",
                                                        "worse",
                                                        "better")),
                  bladder_long_term = factor(dt1$`Long-term bladder`,
                                             levels = 0:1,
                                             labels = c("No",
                                                        "Yes")),
                  bladder_long_term_change = factor(tolower(dt1$`Long-term bladder change`),
                                                    levels = c("unchanged",
                                                               "worse",
                                                               "better")),
                  
                  #mMCs
                  mmcs_pre = factor(dt1$`pre mMCs`,
                                    levels = 1:5),
                  mmcs_3m = factor(dt1$`3m mMCs`,
                                    levels = 1:5),
                  mmcs_3m_change = factor(tolower(dt1$`3m mMCs change`),
                                          levels = c("unchanged",
                                                     "worse",
                                                     "better")),
                  mmcs_long_term  = factor(dt1$`Long-term mMCs`,
                                    levels = 1:5),
                  mmcs_long_term_change = factor(tolower(dt1$`Long-term mMCs change`),
                                                 levels = c("unchanged",
                                                            "worse",
                                                            "better")),
                  
                  # ASIA IS
                  asia_is_pre = factor(dt1$`pre ASIA IS`,
                                       levels = LETTERS[1:5]),
                  asia_is_3m = factor(dt1$`3m ASIA IS`,
                                       levels = LETTERS[1:5]),
                  asia_is_3m_change = factor(tolower(dt1$`3m ASIA IS change`),
                                             levels = c("unchanged",
                                                        "worse",
                                                        "better")),
                  
                  asia_is_long_term = factor(dt1$`Long-term ASIA IS`,
                                       levels = LETTERS[1:5]),
                  asia_is_long_term_change = factor(tolower(dt1$`Long-term ASIA IS change`),
                                                    levels = c("unchanged",
                                                               "worse",
                                                               "better")),
                  
                  # BCR
                  bcr_baseline = factor(tolower(dt1$baseline_BCR),
                                        levels = c("not detected",
                                                   "no response (sph sin)",
                                                   "normal",
                                                   "n/a")),
                  bcr_intraoperative = factor(tolower(dt1$Intraoperative_BCR),
                                              levels = c("unchanged",
                                                         "amplitude decrease (sph dx)",
                                                         "not detected",
                                                         "n/a")),
                  
                  spinal_segment = factor(dt1$`spinal segment`), # group segments into sections
                  surgical_radicality = factor(dt1$`surgical radicality`),
                  # dt1$`neurophysiology used` # all = 1
                  adjuvent_radio = factor(dt1$`adjuvent radio`,
                                          levels = 0:1,
                                          labels = c("No",
                                                     "Yes")),
                  adjuvant_chemo = factor(dt1$`adjuvant chemo`,
                                          levels = 0:1,
                                          labels = c("No",
                                                     "Yes")),
                  
                  # Motor evoked potentials
                  # baseline_MEP_right_hand = factor(tolower(dt1$baseline_MEP_right_hand)),
                  # baseline_MEP_left_hand = factor(tolower(dt1$baseline_MEP_left_hand)),
                  baseline_MEP_right_leg = factor(tolower(dt1$baseline_MEP_right_leg)),
                  baseline_MEP_left_leg = factor(tolower(dt1$baseline_MEP_left_leg)),

                  Intraoperative_MEP_right_hand = factor(tolower(dt1$Intraoperative_MEP_right_hand)),
                  Intraoperative_MEP_left_hand = factor(tolower(dt1$Intraoperative_MEP_left_hand)),
                  Intraoperative_MEP_right_leg = factor(tolower(dt1$Intraoperative_MEP_right_leg)),
                  Intraoperative_MEP_left_leg = factor(tolower(dt1$Intraoperative_MEP_left_leg)),
                 
                  # Sensory evoked potentials
                  baseline_SEP_right_hand = factor(tolower(dt1$baseline_SEP_right_hand)),
                  baseline_SEP_left_hand = factor(tolower(dt1$baseline_SEP_left_hand)),
                  baseline_SEP_right_leg = factor(tolower(dt1$baseline_SEP_right_leg)),
                  baseline_SEP_left_leg = factor(tolower(dt1$baseline_SEP_left_leg)),

                  Intraoperative_SEP_right_hand = factor(tolower(dt1$Intraoperative_SEP_right_hand),
                                                         levels = c("unchanged",
                                                                    "absent from the start of the surgery",
                                                                    "amplitude decrease",
                                                                    "amplitude decrease, prolonged latency",
                                                                    "loss of response",
                                                                    "n/a")),
                  Intraoperative_SEP_left_hand = factor(tolower(dt1$Intraoperative_SEP_left_hand)),
                  Intraoperative_SEP_right_foot = factor(tolower(dt1$Intraoperative_SEP_right_foot)),
                  Intraoperative_SEP_left_foot = factor(tolower(dt1$Intraoperative_SEP_left_foot))
)
```

# Table 1: D-wave Distal Baseline
```{r}
t1 <- table1(~ .| d_wave_distal_baseline,
             data = dt2[, -c("study_id",
                             "op_date",
                             "surgery_date",
                             "long_term_followup_date",
                             "death_date")])

# formatters::export_as_rtf(x = t1,
#                           file = "t1.rtf")

t1
```

# Table 2: D-wave Distal Intraoperative
```{r}
table1(~ .| d_wave_distal_intraoperative,
       data = dt2[, -c("study_id",
                       "surgery_date",
                       "death_date",
                       "long_term_followup_date",
                       "op_date")])
```

```{r}
addmargins(table(motor_pre = dt2$motor_pre,
                 motor_3m_change = dt2$motor_3m_change,
                 motor_long_term_change = dt2$motor_long_term_change))
```

# Figures
## Figure 1: mMCs Scores
```{r}
dt_f1 <- melt.data.table(data = dt2[, c("study_id",
                                        "d_wave_distal_baseline",
                                        "mmcs_pre",
                                        "mmcs_3m",
                                        "mmcs_long_term")],
                         id.vars = 1:2,
                         measure.vars = 3:5,
                         variable.name = "Followup",
                         value.name = "mMCs") 

dt_f1$study_id <- factor(dt_f1$study_id)

dt_f1$Followup <- factor(dt_f1$Followup,
                         levels = c("mmcs_pre",
                                    "mmcs_3m",
                                    "mmcs_long_term"),
                         labels = c("Pre-Op",
                                    "3 Months",
                                    "Long Term"))
```

```{r,fig.width=6,fig.height=3}
ggplot(dt_f1,
       aes(x = Followup,
           y = mMCs,
           group = study_id,
           color = study_id)) +
  facet_wrap(~ d_wave_distal_baseline) +
  geom_line(position = position_dodge(0.3)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1),
        legend.position = "none")
```

## Figure 1: ASIA IS Scores
```{r}
dt_f2 <- melt.data.table(data = dt2[, c("study_id",
                                        "d_wave_distal_baseline",
                                        "asia_is_pre",
                                        "asia_is_3m",
                                        "asia_is_long_term")],
                         id.vars = 1:2,
                         measure.vars = 3:5,
                         variable.name = "Followup",
                         value.name = "ASIA IS") 

dt_f2$study_id <- factor(dt_f2$study_id)

dt_f2$Followup <- factor(dt_f2$Followup,
                         levels = c("asia_is_pre",
                                    "asia_is_3m",
                                    "asia_is_long_term"),
                         labels = c("Pre-Op",
                                    "3 Months",
                                    "Long Term"))

dt_f2$`ASIA IS` <- factor(dt_f2$`ASIA IS`,
                          levels = LETTERS[1:5])
```

```{r,fig.width=6,fig.height=3}
ggplot(dt_f2,
       aes(x = Followup,
           y = `ASIA IS`,
           group = study_id,
           color = study_id)) +
  facet_wrap(~ d_wave_distal_baseline) +
  geom_line(position = position_dodge(0.3)) +
  scale_y_discrete(drop = FALSE) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1),
        legend.position = "none")
```

# Tests
Correlation of intraoperative changes in SEP, MEP and D-wave vs. Clinical Variables at 3 months and long-term


Motor Evoked Potentials (MEP)

## Intraoperative SEP Feet vs. Sensory Change at 3 Months
Sensory Evoked Potentials (SEP)

### Merge SEP right and left feet
```{r}
addmargins(table(Intraoperative_SEP_right_foot = dt2$Intraoperative_SEP_right_foot,
                 motor_3m = dt2$sensory_3m_change))

addmargins(table(Intraoperative_SEP_left_hand = dt2$Intraoperative_SEP_left_foot,
                 motor_3m = dt2$sensory_3m_change))
```

```{r}
unique(tolower(dt2$Intraoperative_SEP_right_foot))

dt2$intraop_sep_rf <- unlist(lapply(X = substr(x = tolower(dt2$Intraoperative_SEP_right_foot),
                                                 start = 1,
                                                 stop = 3),
                                      FUN = function(a) {switch(a,
                                                                "unc" = "unchanged",
                                                                "abs" = "unchanged",
                                                                "los" = "loss of response",
                                                                "amp" = "amplitude decrease",
                                                                "non distinct respone")}))
dt2$intraop_sep_rf <- factor(dt2$intraop_sep_rf,
                             levels = c("unchanged",
                                        "loss of response",
                                        "amplitude decrease",
                                        "non distinct respone"))
table(dt2$intraop_sep_rf)
```


```{r}
unique(tolower(dt2$Intraoperative_SEP_left_foot))

dt2$intraop_sep_lf <- unlist(lapply(X = substr(x = tolower(dt2$Intraoperative_SEP_left_foot),
                                                 start = 1,
                                                 stop = 3),
                                      FUN = function(a) {switch(a,
                                                                "unc" = "unchanged",
                                                                "lat" = "unchanged",
                                                                "abs" = "unchanged",
                                                                "los" = "loss of response",
                                                                "amp" = "amplitude decrease",
                                                                "non distinct respone")}))
dt2$intraop_sep_lf <- factor(dt2$intraop_sep_lf,
                             levels = c("unchanged",
                                        "loss of response",
                                        "amplitude decrease",
                                        "non distinct respone"))
table(dt2$intraop_sep_lf)
```

```{r}
addmargins(table(rf = dt2$intraop_sep_rf,
                 lf = dt2$intraop_sep_lf))

dt2$intraop_sep_feet <- "unchanged"

dt2$intraop_sep_feet[dt2$intraop_sep_rf == "loss of response" |
                       dt2$intraop_sep_lf == "loss of response"] <- "loss of response"

dt2$intraop_sep_feet[dt2$intraop_sep_rf == "amplitude decrease" |
                       dt2$intraop_sep_lf == "amplitude decrease"] <- "amplitude decrease"

dt2$intraop_sep_feet[dt2$intraop_sep_rf == "non distinct respone" |
                       dt2$intraop_sep_lf == "non distinct respone"] <- "non distinct respone"
dt2$intraop_sep_feet <- factor(dt2$intraop_sep_feet,
                               levels = c("unchanged",
                                          "loss of response",
                                          "amplitude decrease",
                                          "non distinct respone"))

addmargins(table(dt2$intraop_sep_feet))
```

```{r}
s11 <- summary(glm((sensory_3m_change == "worse") ~ dt2$intraop_sep_feet,
                   data = dt2,
                   family = "binomial"))
s11

out11 <- data.table(Y = "Motor Worse at 3 Months",
                  X = "Intraoperative SEP Feet",
                  Comparisons = paste(levels(dt2$intraop_sep_feet)[-1],
                                      "vs. unchanged"),
                  OR = round(exp(s11$coefficients[-1, 1]),
                             2))
print("Odds of getting worst motor function at 3 months followup if the Intraop SEP Right Hand had a status XXX vs unchanged")
out11
```

```{r}
s11 <- summary(glm((sensory_3m_change == "better") ~ dt2$intraop_sep_feet,
                   data = dt2,
                   family = "binomial"))
s11

out11 <- data.table(Y = "Motor Worse at 3 Months",
                  X = "Intraoperative SEP Feet",
                  Comparisons = paste(levels(dt2$intraop_sep_feet)[-1],
                                      "vs. unchanged"),
                  OR = round(exp(s11$coefficients[-1, 1]),
                             2))
print("Odds of getting worst motor function at 3 months followup if the Intraop SEP Right Hand had a status XXX vs unchanged")
out11
```


```{r}

# Same as:
22/(62 - 22)
0.9/(1 - 0.9)
(0.9/(1 - 0.9))/(22/(62 - 22))


s12 <- summary(glm((motor_3m_change == "better") ~ Intraoperative_SEP_right_hand,
                   data = dt2,
                   family = "binomial"))
summary(s12)

out12 <- data.table(Y = "Motor Better at 3 Months",
                    X = "Intraoperative SEP Feet",
                    Comparisons = paste(levels(dt2$Intraoperative_SEP_right_hand)[-1],
                                        "vs. unchanged"),
                    OR = round(exp(s12$coefficients[-1, 1]),
                               2))
out12


m13 <- glm((motor_3m_change == "unchanged") ~ Intraoperative_SEP_right_hand,
          data = dt2,
          family = "binomial")
summary(m13)
```



# Save data
```{r}
write.csv(dt2,
          file = "gayane_data_10082023.CSV",
          row.names = TRUE)
```

# Session
```{r}
sessionInfo()
```